<!---
# Part of RHRT: R package to assess Heart Rate Turbulence from RR interval data 
# Copyright (C) 2021 Valeria Blesius

# RHRT is free software: you can redistribute it and/or modify
# it under the terms of the GNU General Public License as published by
# the Free Software Foundation, version 2 only.

# RHRT is distributed in the hope that it will be useful,
# but WITHOUT ANY WARRANTY; without even the implied warranty of
# MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
# GNU General Public License for more details.

# You should have received a copy of the GNU General Public License
# along with RHRT.  If not, see <https://www.gnu.org/licenses/>.
-->

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RHRT

RHRT provides tools for a Heart Rate Turbulence analysis of RR interval data. It can either find the ventricular premature complexes (VPCs) via a set of filter rules or can use annotation data to only check the beats with the correct annotation. VPC snippets are filtered for validity and HRT parameters calculated.

In addition to standard calculation methods the package allows to modify the filter and calculation parameters. It is therefore not only helpful to identify HRT classes of measurements for risk assessment but also for assessment of the methodology itself.

For more information please check the vignette of the package.
For more information about HRT have a look into the [original publication by Schmidt et al.](https://doi.org/10.1016/S0140-6736(98)08428-1) or our [review](https://doi.org/10.1088/1361-6579/ab98b3) with focus on the methodology.

## Installation

You can install the released version of RHRT from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("RHRT")
```

And the development version from [GitHub](https://github.com/) with:

``` r
install.packages("devtools") # if not already done
devtools::install_github("VBlesius/RHRT/RHRT")
```
## Example

The general workflow of RHRT is to scan your interval data for HRT and check the results via HRT class and plot:

```{r example, fig.width=7, fig.height=4}
library(RHRT)
## scan your interval data and save the results as an HRTList
hrtl <- vectorToHRT(testdataLong)
## get the HRT class of your data
getResults(hrtl, type = "class")
## have a look at the data and the parameters
plot(hrtl)
```

## Data

Data to test the package can be found on [Physionet](https://physionet.org/). Via the [WFDB Toolkit](https://physionet.org/content/wfdb/10.6.2/) ECG data can be downloaded and/or converted, for example:

```{bash, eval = FALSE}
ann2rr -r chf2db/chf201 -a ecg -i s3 -w > ~/some/path/chf201.csv
```

Then load the data and use RHRT to find VPCSs:
``` {r, eval = FALSE}
chf201 <- read.table("~/some/path/chf201.csv")
hrtl <- RHRT::vectorToHRT(chf201[[1]]*1000, ann = chf201[[2]])
```

More example workflows can be found in the vignette.